Data Specialization for a Staged Lambda Calculus

### 2015-02-01 paper/splitting.tex
various edits to spliting section

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as it ability to form
mixed-stage abstractions.
[Sentence about how we actual want this in separated monostage form.]
In this section, we introduce a {\em splitting transformation} whose job 
is to translate \lang\ terms into a form where the stages are 
seperated {\em at the top level}.

Recall that $\colmix {e_1}{A_1}$ represents multistage computations with multistage results,
whereas $\coltwo {e_2}{A_2}$ represents multistage computions with stage \bbtwo\ results.
Similarly, our splitting algorithm has two forms, 
given by the judgements $\splitonesym$ and $\splittwosym$, 
which respectively take these types of \lang\ terms as input.
For the outputs, we introduce new constructs 
to encode multistage values and computations that are syntactically separated at the top level.

@@

Recall that partial values (including residual tables) produced by \bbone-evaluation may
contain content at \emph{both} stages. 
In particular, partial values have the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
For example, the term
\begin{lstlisting}
1`(1+2, next{`2`7 + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}

@@

The exmaple above is masked into,
\begin{lstlisting}
mv((3,()), let yhat=7+5 in ((),yhat))
\end{lstlisting}

@@

Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
To represent this in a separated form, we use the construct $\pipeM c l r$, 
where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
This ``multi-output pipeline'' can be partially evaluated with the following rule:
\[
\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
where $i$ becomes the immediate result of the masked value,
and where $b$ is plugged into the open variable of the resumer.
We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.

@@

Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
splitting stage~\bbtwo\ terms in \lang\ assumes a simpler form than that of stage~\bbone\ term splitting. 
Specifically, splitting a stage~\bbtwo\ term $e$ in \lang\ generates a precomputation term $p$
(performing all stage~\bbone\ subcomputations in $e$) and a
resumer term $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
$e$ reduces to the value~$v$, $\letin l p r$ reduces to an identical value in \langmono.
Note that in contrast to the combined term produced during stage~\bbone\ splitting,
the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
to compute the boundary value expected by the resumer.

@@

A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
\ur{This last point seems important.  It would be important to come
  back to it.}

>>>>

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as it ability to form
mixed-stage abstractions.
To achieve this, \lang\ has three classes of multistage construct:
\begin{itemize}
\item terms $\colmix {e_1}{A_1}$ to represent multistage computations with multistage results,
\item terms $\coltwo {e_2}{A_2}$ to represent multistage computions with purely stage \bbtwo\ results, and
\item patial values $[\xi; v]$ to represent computations which have produced stage \bbone\ results,
but for which some stage \bbtwo\ computation remains.
\end{itemize}

Our goal in this section will be to, for each class above, 
show how to translate members of that class into
an equivalent form where the stages are separated at the top level.

More formally, we introduce three new constructs 
(\texttt{mv}, \texttt{pipeM}, \texttt{pipeS}) 
to represent the separated form of each class, 
and three judgements ($\vsplito$,$\splitonesym$,$\splittwosym$)
mapping interleaved forms to separated forms.
In summary:

\begin{center}
\begin{tabular} {|c|c|c|c|}
\hline
Class
	& \begin{tabular}[x]{@{}c@{}}Interleaved\\Form\end{tabular} 
	& \begin{tabular}[x]{@{}c@{}}Separated\\Form\end{tabular} 
	& \begin{tabular}[x]{@{}c@{}}Splitting\\Relation\end{tabular} 
	\\ \hline
Partial Value
	& $[\xi; v]$
	& $\mval i r$
	& $\vsplito$
	\\ \hline
\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Outputs at \bbone,\bbtwo)\end{tabular} 
	& $\colmix e A$
	& $\pipeM c l r$
	& $\splitonesym$
	\\ \hline
\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Output at \bbtwo)\end{tabular} 
	& $\coltwo e A$
	& $\pipeS p l r$
	& $\splittwosym$
	\\ \hline
\end{tabular}
\end{center}

@@

Recall that partial values are multistage terms with the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.

For example, the term
\begin{lstlisting}
1`(1+2, next{`2`7 + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}

@@

The example above is masked into,
\begin{lstlisting}
mv((3,()), let yhat=7+5 in ((),yhat))
\end{lstlisting}

@@

Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
To represent this in a separated form, we use the ``multi-output pipeline'' construct $\pipeM c l r$, 
where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
This can be partially evaluated (into a masked value) with the following rule:
\[
\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
where $i$ becomes the immediate result of the masked value,
and where $b$ is plugged into the open variable of the resumer.
We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.

@@

Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
stage~\bbtwo\ term splitting assumes a simpler form than does stage~\bbone\ term splitting. 
Specifically, splitting a term $\coltwo e A$ in \lang\ generates a ``single-output pipeline''
term $\pipeS p l r$ where $p$ is a monostage {\em precomputation} term 
that performs all of the stage~\bbone\ subcomputations in $e$, and $l.r$ 
is a resumer which that performs all of the stage~\bbtwo\ subcomputations in $e$.
This can be partially evaluated into a residual via the rule,
\[
\infer {\sepredone {\pipeS p l r} {[b/l]r}} {\reduce p b} 
\]
Note that in contrast to the combined term produced during stage~\bbone\ splitting,
the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
to compute the boundary value expected by the resumer.

@@

A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
